---
title: "Second Brexit Referendum"
author: "Marco Basaldella"
date: "June 27, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# install missing packages (if needed) and load them

pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE)
    if(!require(x,character.only = TRUE)) stop("Package not found")
  }
}
pkgTest("jsonlite")
pkgTest("rworldmap")
pkgTest("raster")
pkgTest("rgeos")
pkgTest("rgdal")
pkgTest("ggplot2")
pkgTest("RCurl")
```


After the Brexit referendum, a [petition](https://petition.parliament.uk/petitions/131215) surfaced on the Internet requesting a second referendum. Intuition may tell us that territories where *"Remain"* won may be the ones which are requesting a second petition, but is this the case?

Luckily, the UK Parliament website offers all the petition data in JSON format available for download, so we can freely analyze them with RStudio.

To start, we simply download the content of the petition data into a variable:

```{r}
json_data <- fromJSON("https://petition.parliament.uk/petitions/131215.json",flatten=TRUE)
```

If we analyze the output with `str(json_data)` (not shown), we can see that we have the petition data divided by country and by [parliament constiuencies](https://en.wikipedia.org/wiki/List_of_United_Kingdom_Parliament_constituencies). 

First of all, we start by looking who in the world would love to keep Britain in the UK. So, we keep only the signatures country data:

```{r}
# filter the petitions by country
sgn_by_country = json_data$data$attributes$signatures_by_country

```

Then, we look at the top countries who want Britain in the EU. We sort the array and display the top 20 countries using ```head```:

```{r}
sgn_by_country = sgn_by_country[order(-sgn_by_country$signature_count),]
head(sgn_by_country,n=20)
```

Ooh, so it seems that French don't hate Britons so much! The number of signatures from Gibraltar is also very interesting, considering that the population of the territory is just about [35,000](https://en.wikipedia.org/wiki/Gibraltar).

So, to have a clearer picture, let's put this data on a [beautiful world map](https://github.com/AndySouth/rworldmap)!

```{r, message=FALSE,warning=FALSE, fig.width=12, fig.height=8}
sgn_by_country$log_signature_count = log10(sgn_by_country$signature_count)
dataMap <- joinCountryData2Map(sgn_by_country,joinCode="ISO2",nameJoinColumn="code")
mapData = mapCountryData(dataMap, nameColumnToPlot="signature_count",mapTitle="Second Brexit Referendum",catMethod=10^(0:7),addLegend = FALSE)
do.call( addMapLegendBoxes, c(mapData,title="# signatures",x="bottomleft"))
```

Now let's see at how the petition is popular in Britain. As before, keep the constituencies only from the JSON and look at the top 20.

```{r}
signatures_by_constituency = json_data$data$attributes$signatures_by_constituency
signatures_by_constituency = signatures_by_constituency[order(-signatures_by_constituency$signature_count),]

head(signatures_by_constituency[,c(1,4)],n=20)
```

Now let's do the map. Unfortunately, Nothern Ireland will be exluded from our map, since the geometry data I found doesn't include it.

First, we read the map data and match the name of the constituencies.

```{r}
# load the map
ukMap = readOGR("https://github.com/martinjc/UK-GeoJSON/raw/master/json/electoral/gb/topo_wpc.json","wpc")

# match the names in the map
ind = match(as.character(ukMap@data$id),signatures_by_constituency$ons_code)
ukMap@data$name = signatures_by_constituency$name[ind]
ukMap@data$votes = signatures_by_constituency$signature_count[ind]
```

Then, we prepare the dataframe for plotting:

```{r}
# build the dataframe to print
constituencies.map <- data.frame(id=0:(length(ukMap@data$name)-1),
                                 Constituency=as.character(ukMap@data$name))
plotData = fortify(ukMap)
plotData = merge(plotData,constituencies.map,by="id")
ind = match(plotData$Constituency,as.factor(signatures_by_constituency$name))
plotData$votes = signatures_by_constituency$signature_count[ind]
```

And finally, the maps!

```{r fig.width=8, fig.height=8}
theme_set(theme_minimal())

p = ggplot(data=plotData,
       aes(x=long, y=lat,
           group=group))

p = p + 
  geom_map(data = plotData,
           map = plotData,
           aes(map_id=id, x=long, y=lat, group=group,
               fill=plotData$votes),
           color="white", size=0.001) +
  scale_fill_gradient(
    #guide = "legend",
    breaks = c(0,200,400,800,1600,3200,6400,12800,25600),
    name = "# signatures",
    low='green', high='red', trans="log")

p = p +  
  labs(x="", y="", title="Petition for second Brexit referendum")+ #labels
  theme(axis.ticks.y = element_blank(),axis.text.y = element_blank(), 
        axis.ticks.x = element_blank(),axis.text.x = element_blank(), 
        plot.title = element_text(lineheight=1, face="bold")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  
  
p
```

Hmm, nice. But it's not that useful: we should normalize petition signatures by the number of voters for each region. Thankfully, the UK's electoral commission is very helpful: we can download 2015 electoral data from the UK's [electoral commission](http://www.electoralcommission.org.uk/our-work/our-research/electoral-data) website.

For convenience, I pre-processed and cleaned the data a bit before feeding them into R. Let's see which are the top 20 constituencies by *percentage* of electorate requesting a second referendum:

```{r}
electoralData = read.csv(
  text = getURL("https://raw.githubusercontent.com/basaldella/second-brexit/master/UkConstituencyData2015.csv"),
  header = TRUE,
  stringsAsFactors = FALSE)

electoralData = merge(electoralData,signatures_by_constituency,by="name")
electoralData$signaturePercent = 100* electoralData$signature_count / electoralData$electorate
electoralData = electoralData[order(-electoralData$signaturePercent),]
head(electoralData[,c(1,9)],20)

```

Let's place it again on a beautiful map to make it clearer:

```{r fig.width=8, fig.height=8}
ind = match(plotData$Constituency,as.factor(electoralData$name))
plotData$votesPercent = electoralData$signaturePercent[ind]

p = ggplot(data=plotData,
       aes(x=long, y=lat,
           group=group))

p = p + 
  geom_map(data = plotData,
           map = plotData,
           aes(map_id=id, x=long, y=lat, group=group,
               fill=plotData$votesPercent),
           color="white", size=0.001) +
  scale_fill_gradient(
    #guide = "legend",
    breaks = c(0,5,10,15,20,25),
    name = "% of voters\nrequesting 2nd\nreferendum",
    low='green', high='red')

p = p +  
  labs(x="", y="", title="Petition for second Brexit referendum")+ #labels
  theme(axis.ticks.y = element_blank(),axis.text.y = element_blank(), 
        axis.ticks.x = element_blank(),axis.text.x = element_blank(), 
        plot.title = element_text(lineheight=1, face="bold")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  
  
p
```

Unfortunately, we have some grey areas, where data from different sources doesn't match.

Now we aggregate per region:

```{r}
aggData = electoralData[,c(3,5,8)]
regionalData <-aggregate(data = aggData, electorate~region,FUN=sum, na.rm=TRUE)
regionalData$signaturePercentRegion <-aggregate(data = aggData, signature_count~region,FUN=sum, na.rm=TRUE)$signature_count
regionalData$signaturePercentRegion = 100 * regionalData$signaturePercentRegion / regionalData$electorate

regionalData = regionalData[order(-regionalData$signaturePercentRegion),]
head(regionalData,n=20)
```

Well, it seems to be that the regions where the Leave won are the ones who now are requesting a second referendum the most. Scotland would seem to not care so much - probably, they would just [prefer indipendence](http://www.theguardian.com/politics/live/2016/jun/23/eu-referendum-result-live-counting-leave-remain-brain-in-europe).
The same hold for Nothern Ireland (`NI` in the output), with just 5% of voters requesting the referendum.

Then we plot this data on a map:

```{r fig.width=8, fig.height=8}
# pre-plot 
electoralData = merge(electoralData,regionalData[,c(1,3)],by="region")
ind = match(plotData$Constituency,as.factor(electoralData$name))
plotData$votesPercentRegion = electoralData$signaturePercentRegion[ind]

p = ggplot(data=plotData,
       aes(x=long, y=lat,
           group=group))

p = p + 
  geom_map(data = plotData,
           map = plotData,
           aes(map_id=id, x=long, y=lat, group=group,
               fill=plotData$votesPercentRegion),
           color="white", size=0.001) +
  scale_fill_gradient(
    #guide = "legend",
    #breaks = c(0,5,10,15,20,25),
    name = "% of voters\nrequesting 2nd\nreferendum",
    low='green', high='red')

p = p +  
  labs(x="", y="", title="Petition for second Brexit referendum")+ #labels
  theme(axis.ticks.y = element_blank(),axis.text.y = element_blank(), 
        axis.ticks.x = element_blank(),axis.text.x = element_blank(), 
        plot.title = element_text(lineheight=1, face="bold")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  
  
p

```


Beautiful, eh? This gives a very clear picture on how Englishmen seem to be the ones regretting their decision the most! Anyways, I am by no means a political scientist: I just wanted to have a bit of fun with R, so I'll leave the interpretation of this picture to someone else.

### Credits

I took many ideas and code from [this github repo](https://github.com/kjhealy/uk-elections). Spatial data for the maps comes from [Martin Chorley's UK-GeoJSON project](https://github.com/martinjc/UK-GeoJSON), and it's licensed with a CC-BY 4.0 license. Petition data and electoral body data come from the UK's institutional websites (linked above).

### Disclaimer

Using electoral data to compute these maps is probably not the best choice, because the petition can be signed by anyone, not just people with voting rights.

Moreover, suspicions rose that the petition has been [manipulated](http://www.theguardian.com/politics/2016/jun/26/petition-for-second-eu-referendum-may-have-been-manipulated) by citiziens signing twice, or giving away their postal code. Anyways, the data has been periodically cleaned by the authorities: for example, I own one dump file with thousand of signatures from Vatican City, which have then disappeared.

### Data and source

Source code and processed data for this experiment is available on [github](https://github.com/basaldella/second-brexit).

### License

This notebook is licensed with a [CC-BY 4.0](https://creativecommons.org/licenses/by/4.0/) license. Please attribute by linking back to this file or to the corresponding [github repo](https://github.com/basaldella/second-brexit).

